Engineering the Servo Web Browser Engine using Rust

### 2015-10-22 icse16-servo/rust.tex

  Thread::spawn(move || {
    *data = *data + 1;
  });
  // error: accessing moved value
  print!("{}", data);
}
\end{lstlisting}
  \caption{Code that will not compile because it attempts to access mutable state from two threads.}
  \label{fig:bad-concurrency}
\end{figure}

On the other hand, the immutable value in \figref{fig:shared-concurrency} can be borrowed and shared between multiple threads as long as those threads don't outlive the scope of the data, and even mutable values can be shared as long
as they are owned by a type that preserves the invariant that mutable memory is unaliased, as with the
mutex in \figref{fig:shared-mutable-concurrency}.

\begin{figure}
\begin{lstlisting}
fn main() {
  // An immutably borrowed pointer to a
  // stack-allocated integer
  let data = &1;

  Thread::scoped(|| {
    println!("{}", data);
  });
  print!("{}", data);
}
\end{lstlisting}
  \caption{Safely reading immutable state from two threads.}
  \label{fig:shared-concurrency}
\end{figure}

@@

  Thread::scoped(move || {
    *data2.lock().unwrap() = 1;
  });

@@

With relatively few simple rules, ownership in Rust enables foolproof task parallelism,
but also data parallelism, by partitioning vectors and lending mutable references into properly scoped threads.
Rust's concurrency abstractions are entirely implemented in libraries, and though
many advanced concurrent patterns such as work-stealing~~\cite{blumeofe:multiprogrammed-work-stealing}
cannot be implemented in safe Rust, they can usually be encapsulated in a memory-safe interface.

\subsection{Traits and polymorphism}

Rust supports both static and dynamic parametric polymorphism via \emph{traits},
which build on ideas from type classes and type families in Haskell~\cite{haskell}.

As with Haskell type classes, traits define functions and types that are associated
with all implementing types. Function and types in turn can be parameterized over
generic types that are bounded by traits, as with the trait that defines overloaded
addition in \figref{fig:fn-trait}

Such calls to associated functions are statically dispatched, so optimize well,
but Rust also allows types that implement traits to be packaged with a vtable
so that functions can be resolved dynamically. Because Rust's generics must
perform as well as C++'s inlined templates, performance-sensitive Rust
code tends to prefer the static dispatch of traits-as-typeclasses.
This is true of much of the standard library.

Rust includes several types of closures that are distinguished by whether
they own or borrow their environment, and these are defined as traits.

\begin{figure}
\begin{lstlisting}
pub trait Add<RHS=Self> {
  type Output;

  fn add(self, rhs: RHS) -> Self::Output;
}

fn add_pair<T>(pair: (T, T)) -> T::Output
                             where T: Add {
  pair.0 + pair.1
}
\end{lstlisting}
  \caption{Rust trait with associated type.}
  \label{fig:fn-trait}
\end{figure}

>>>>

  thread::spawn(move || {
    *data = *data + 1;
  });
  // error: accessing moved value
  print!("{}", data);
}
\end{lstlisting}
  \caption{Code that will not compile because it attempts to access mutable state from two threads.}
  \label{fig:bad-concurrency}
\end{figure}

On the other hand, the immutable value in \figref{fig:shared-concurrency} can be borrowed and shared between multiple threads as long as those threads don't outlive the scope of the data, and even mutable values can be shared as long
as they are owned by a type that preserves the invariant that mutable memory is unaliased, as with the
mutex in \figref{fig:shared-mutable-concurrency}. It is also possible to share stack-local data across appropriately scoped threads safely without extra cost.\footnote{\url{https://github.com/Kimundi/scoped-threadpool-rs}}

\begin{figure}
\begin{lstlisting}
fn main() {
  // An immutable shared pointer
  // (atomically reference counted)
  let data = Arc::new(1);

  thread::spawn(|| {
    println!("{}", *data);
  });
  print!("{}", *data);
}
\end{lstlisting}
  \caption{Safely reading immutable state from two threads.}
  \label{fig:shared-concurrency}
\end{figure}

@@

  thread::spawn(move || {
    *data2.lock().unwrap() = 1;
  });

@@

With relatively few simple rules, ownership in Rust enables foolproof task parallelism,
but also data parallelism, by partitioning vectors and lending mutable references across threads.
Rust's concurrency abstractions are entirely implemented in libraries, and though
many advanced concurrent patterns such as work-stealing~~\cite{blumeofe:multiprogrammed-work-stealing}
cannot be implemented in safe Rust, they can usually be encapsulated in a memory-safe interface.
